`mathml`: Translate R Expressions to MathML and LaTeX

### 2023-06-20 rj/mathml.Rmd
small changes

The R\ extension of the markdown language [@Xie2018;rmarkdown] enables
reproducible statistical reports with nice typesetting in HTML, Microsoft Word,
and LaTeX. Since recently [@R, version 4.2], R\'s manual pages include support
for mathematical expressions [@Sarkar2022;@Viechtbauer2022], which already is a
big improvement. However, except for special cases such as regression
models [@equatiomatic] and R's own plotmath annotation, rules for the mapping of
built-in language elements to their mathematical representation are still
lacking. So far, R\ expressions such as `pbinom(k, N, p)` are printed as they
are; pretty mathematical formulae such as \(P_{\mathrm{Bi}}(X \le k; N, p)\)
require explicit LaTeX commands
like `P_{\mathrm{Bi}}\left(X \le k; N, p\right)`. Except for very basic use
cases, these commands are tedious to type in and their source code is hard to
read.

The present R\ package defines a set of rules for the automatic translation of
R\ expressions to mathematical output in R\ Markdown documents [@Xie2020] and
Shiny apps [@Chang2022]. The translation is done by an embedded Prolog
interpreter that maps nested expressions recursively to MathML and
LaTeX/MathJax, respectively. User-defined hooks enable to extend the set of
rules, for example, to represent specific R\ elements by custom mathematical
signs.

The main feature of the package is that the same R\ expressions and equations
can be used for both mathematical typesetting and calculations. This saves time
and potentially reduces mistakes, as will be illustrated below. Readers should
have basic knowledge of knitr and R\ Markdown to be able to follow this
article [@knitr;@rmarkdown]. To extend and customize the package, some basic
knowledge of Prolog is needed.

@@

Similar to other high-level programming languages, R is homoiconic, that is,
R\ commands (i.e., R\ "calls") are, themselves, symbolic data structures that
can be created, parsed and modified. Because the default response of the
R\ interpreter is to evaluate a call and return its result, this property is not
transparent to the general user. There exist, however, a number of built-in
R\ functions (e.g., `quote()`, `call()` etc.) that allow the user to create R\ calls
which can be stored in regular variables and then, for example, be evaluated at
a later stage or in a specific environment [@Wickham2019]. The present package
includes a set of rules that translate such calls to a mathematical
representation in MathML and LaTeX. For a first illustration of the \pkg{mathml}
package, we consider the binomial probability.

@@

As is seen from the output, the variable `term` is not assigned the result of
the calculation, but an R\ call
[see, e.g., @Wickham2019, for details on "non-standard evaluation"]. Such a call
can eventually be evaluated with `eval()`,

@@

For self-written functions, matters are a bit more complicated. For a function
such as `g <- function(...) ...`, the name _g_ is not transparent to R, because
only the function body is represented. We can still display functions in the
form `head(x) = body` if we embed the object to be shown into a
call `"<-"(head, body)`.

@@

This package allows R\ to render its terms in pretty mathematical equations. It
extends the current features of R\ and existing packages for displaying
mathematical formulas in R\ [@murrell2000;@allaire2018]), but most
importantly, \pkg{mathml} bridges the gap between computational needs,
presentation of results, and their reproducibility. The package supports both
MathML and LaTeX/MathJax for use in R\ Markdown documents, presentations and
ShinyApp webpages.

@@

One limitation of the package is the lack of a convenient way to insert line
breaks. This is mostly due to lacking support by MathML and LaTeX renderers. 
For example, in its current stage, the LaTeX package `breqn` [@breqn] is mostly
a proof of concept. Moreover, \pkg{mathml} is one-way road, that is, translation
from LaTeX or HTML back to R is not possible [see @latex2r, for an example].

>>>>

The R\ extension of the markdown language [@Xie2018;@rmarkdown] enables
reproducible statistical reports with nice typesetting in HTML, Microsoft Word,
and LaTeX. Moreover, since recently [@R, version 4.2], R\'s manual pages 
include support for mathematical expressions [@Sarkar2022;@Viechtbauer2022],
which already is a big improvement. However, except for special cases 
such as regression models [@equatiomatic] and R's own plotmath annotation, 
rules for the mapping of built-in language elements to their mathematical 
representation are still lacking. So far, R\ expressions such 
as `pbinom(k, N, p)` are printed as they are and pretty mathematical formulae 
such as \(P_{\mathrm{Bi}}(X \le k; N, p)\) require explicit LaTeX commands
like `P_{\mathrm{Bi}}\left(X \le k; N, p\right)`. Except for very basic use
cases, these commands are tedious to type in and their source code is hard to
read.

The present R\ package defines a set of rules for the automatic translation of
R\ expressions to mathematical output in R\ Markdown documents [@Xie2020] and
Shiny Apps [@Chang2022]. The translation is done by an embedded Prolog
interpreter that maps nested expressions recursively to MathML and
LaTeX/MathJax, respectively. User-defined hooks enable to extend the set of
rules, for example, to represent specific R\ elements by custom mathematical
signs.

The main feature of the package is that the same R\ expressions and equations
can be used for both mathematical typesetting and calculations. This saves time
and potentially reduces mistakes, as will be illustrated below. Readers should
have basic knowledge of knitr and R\ Markdown to be able to follow this
article [@knitr;@rmarkdown], while to extend and customize the package, 
some basic knowledge of Prolog is needed.

@@

Similar to other high-level programming languages, R is homoiconic, that is,
R\ commands (i.e., R\ "calls") are, themselves, symbolic data structures that
can be created, parsed and modified. Because the default response of the
R\ interpreter is to evaluate a call and return its result, this property is not
transparent to the general user. There exists, however, a number of built-in
R\ functions (e.g., `quote()`, `call()` etc.) that allow the user to create R\ calls
which can be stored in regular variables and then, for example, evaluated at
a later stage or in a specific environment [@Wickham2019]. The present package
includes a set of rules that translate such calls to a mathematical
representation in MathML and LaTeX. For a first illustration of the \pkg{mathml}
package, we consider the binomial probability.

@@

As it can be seen from the output, to the variable `term` is not assigned the 
result of the calculation, yet an R\ call [see, e.g., @Wickham2019, for details 
on "non-standard evaluation"], which can eventually be evaluated with `eval()`.

@@

For self-written functions, the matter is somewhat more complicated. For a function
such as `g <- function(...) ...`, the name _g_ is not transparent to R, because
only the function body is represented. We can still display functions in the
form `head(x) = body` if we embed the object to be shown into a
call `"<-"(head, body)`.

@@

This package allows R\ to render its terms in pretty mathematical equations. It
extends the current features of R\ and existing packages for displaying
mathematical formulas in R\ [@murrell2000;@allaire2018]), but most
importantly, \pkg{mathml} bridges the gap between computational needs,
presentation of results, and their reproducibility. The package supports both
MathML and LaTeX/MathJax for use in R\ Markdown documents, presentations and
Shiny App webpages.

@@

One limitation of the package is the lack of a convenient way to insert line
breaks. This is mostly due to lacking support by MathML and LaTeX renderers. 
For example, in its current stage, the LaTeX package `breqn` [@breqn] is mostly
a proof of concept. Moreover, \pkg{mathml} is one-way road, that is, it is not 
possible to translate from LaTeX or HTML back to R [see @latex2r, for an example].
